---
title: "Lecture_7"
author: "Matthew Reyers"
date: "October 9, 2019"
output: html_document
---

# 1. Complete the default fit for the gam

```{r, message=FALSE, warning=FALSE}
# Lecture 7 work, dealing with GAMs and PPR

library(glmnet)
library(tidyverse)
library(mgcv)
library(GGally)
library(gam)

prostate <- read.table("Prostate.csv", header=TRUE, sep=",", na.strings=" ")

set.seed(120401002)
prostate$set <- ifelse(runif(n=nrow(prostate))>0.5, yes=2, no=1)

prostate_set1 <- prostate %>%
  filter(set %in% 1)

failed_model <- gam(data = prostate_set1, lpsa ~ s(lcavol) + s(lweight) + s(age) + s(lbph) + svi + s(lcp) +
      gleason + s(pgg45))

ggpairs(prostate_set1 %>% select(-ID, -train, -set))

# Pairs scatterplot seems to actually suggest that lbph is less useful than lcp, but I will remove lcp for now

fixed_model <- mgcv::gam(data = prostate_set1, lpsa ~ s(lcavol) + s(lweight) + s(age) + s(lbph) + svi +
                     gleason + s(pgg45))


```

# a) Report the p-values of the model, discussing the relative importance of the variables.

```{r}

# a) Report p-values, discuss relative importance
# The p-values are accessed like this via mgcv::gam. gam::gam implementations do this differently
model_sum <- summary(fixed_model)
model_sum$p.pv
model_sum$s.pv

```

The p-values must be accessed using two different calls, one for linear terms and one for smoothed terms. Overall, the most important terms by p-value tend to be the smoothed lcavol, lweight, and age in that order. Only the linear term for svi seems to be roughly as important as the 3 smoothed terms above.

# b) To get a more parsimonious model, do some model fitting via backwards elimination. Remove variables based on p-value.
```{r}
# Testing mgcv::gam vs. gam::gam (default)
# Tom uses mgcv so it seems to be the case that this approach is more valid
prostate_set1_reduced <- prostate_set1 %>%
  dplyr::select(-ID, -train, -set)
bic_values_mgcv <- rep(0, 8)
# How can I define a function for this?
# I want to see if I can generalize this process
# Needs to be able to make a spline of some terms and not of others
# Complicated
# For now, do it in the non-scaleable manual way with p-values
base_model <- mgcv::gam(data = prostate_set1_reduced, lpsa ~ s(lcavol) + s(lweight) + s(age) + s(lbph) + svi +
                     gleason + s(pgg45))
bic_values_mgcv[1] <- extractAIC(base_model, k = log(nrow(prostate_set1_reduced)))[2]

# The worst in the base_model is pgg45, replace this with lcp
base_model_lcp <- mgcv::gam(data = prostate_set1_reduced, lpsa ~ s(lcavol) + s(lweight) + s(age) + s(lbph) + svi +
                     gleason + s(lcp))
bic_values_mgcv[2] <- extractAIC(base_model_lcp, k = log(nrow(prostate_set1_reduced)))[2]

# Removing variables, so the highest p-value variable (excluding intercept) is lbph
reduced_model_1 <- mgcv::gam(data = prostate_set1_reduced, lpsa ~ s(lcavol) + s(lweight) + s(age) + svi + 
                       gleason + s(lcp))
bic_values_mgcv[3] <- extractAIC(reduced_model_1, k = log(nrow(prostate_set1_reduced)))[2]

# now lcp
reduced_model_2 <- mgcv::gam(data = prostate_set1_reduced, lpsa ~ s(lcavol) + s(lweight) + s(age) + svi + 
                       gleason)
bic_values_mgcv[4] <- extractAIC(reduced_model_2, k = log(nrow(prostate_set1_reduced)))[2]

# now age
reduced_model_3 <- mgcv::gam(data = prostate_set1_reduced, lpsa ~ s(lcavol) + s(lweight) + svi + gleason)
bic_values_mgcv[5] <- extractAIC(reduced_model_3, k = log(nrow(prostate_set1_reduced)))[2]

# gleason has the highest p-value so that will be removed next
reduced_model_4 <- mgcv::gam(data = prostate_set1_reduced, lpsa ~ s(lcavol) + svi + s(lweight))
bic_values_mgcv[6] <- extractAIC(reduced_model_4, k = log(nrow(prostate_set1_reduced)))[2]

# svi has the highest p-value so that will be removed next
reduced_model_5 <- mgcv::gam(data = prostate_set1_reduced, lpsa ~ s(lcavol) + s(lweight))
bic_values_mgcv[7] <- extractAIC(reduced_model_5, k = log(nrow(prostate_set1_reduced)))[2]

# now lweight
reduced_model_6 <- mgcv::gam(data = prostate_set1_reduced, lpsa ~ s(lweight))
bic_values_mgcv[8] <- extractAIC(reduced_model_6, k = log(nrow(prostate_set1_reduced)))[2]

plot(bic_values_mgcv)


```

```{r}
# Check if there are differences between mgcv and gam
# There are, I believe Tom wants us to use mgcv


# b) Want a more parsimonious model, do some fitting and describe the path through the models. Use BIC and
# effective DF in the answer. Backwards elimination process. 

prostate_set1_reduced <- prostate_set1 %>%
  select(-ID, -train, -set)
bic_values <- rep(0, 8)
# How can I define a function for this?
# I want to see if I can generalize this process
# Needs to be able to make a spline of some terms and not of others
# Complicated
# For now, do it in the non-scaleable manual way with p-values
base_model <- gam(data = prostate_set1_reduced, lpsa ~ s(lcavol) + s(lweight) + s(age) + s(lbph) + svi +
                     gleason + s(pgg45))
bic_values[1] <- extractAIC(base_model, k = log(nrow(prostate_set1_reduced)))[2]

# The worst in the base_model is pgg45, replace this with lcp
base_model_lcp <- gam(data = prostate_set1_reduced, lpsa ~ s(lcavol) + s(lweight) + s(age) + s(lbph) + svi +
                     gleason + s(lcp))
bic_values[2] <- extractAIC(base_model_lcp, k = log(nrow(prostate_set1_reduced)))[2]

# age has the highest p value so that will be removed next
reduced_model_1 <- gam(data = prostate_set1_reduced, lpsa ~ s(lcavol) + s(lweight) + s(lbph) + svi + 
                       gleason + s(lcp))
bic_values[3] <- extractAIC(reduced_model_1, k = log(nrow(prostate_set1_reduced)))[2]

# lbph has the highest p value so that will be removed next
reduced_model_2 <- gam(data = prostate_set1_reduced, lpsa ~ s(lcavol) + s(lweight) + svi + 
                       gleason + s(lcp))
bic_values[4] <- extractAIC(reduced_model_2, k = log(nrow(prostate_set1_reduced)))[2]

# Assuming that the highest p-value to be removed is just the highest value that occurs in either
# the ANOVA for Parametric or Nonparametric effects, lweight is removed next
reduced_model_3 <- gam(data = prostate_set1_reduced, lpsa ~ s(lcavol) + svi +
                       gleason + s(lcp))
bic_values[5] <- extractAIC(reduced_model_3, k = log(nrow(prostate_set1_reduced)))[2]

# # # Continue # # #
# gleason has the highest p-value so that will be removed next
reduced_model_4 <- gam(data = prostate_set1_reduced, lpsa ~ s(lcavol) + svi + s(lcp))
bic_values[6] <- extractAIC(reduced_model_4, k = log(nrow(prostate_set1_reduced)))[2]

# lcp has the highest p-value so that will be removed next
reduced_model_5 <- gam(data = prostate_set1_reduced, lpsa ~ s(lcavol) + svi)
bic_values[7] <- extractAIC(reduced_model_5, k = log(nrow(prostate_set1_reduced)))[2]

# Again, despite being a lower p-value in the Parametric effects, the largest p-value is lcavol when considered as a Nonparametric effect so it will be removed next
reduced_model_6 <-  gam(data = prostate_set1_reduced, lpsa ~ svi)
bic_values[8] <- extractAIC(reduced_model_6, k = log(nrow(prostate_set1_reduced)))[2]

plot(bic_values)

# A more reasonable way to do things, using the stepGAIC function from gamlss
# This makes use of the pgg45 variable being the worst predictor and therefore being removed for lcp
library(gamlss)
test_mod <- gamlss(data = prostate_set1_reduced, lpsa ~ s(lcavol) + s(lweight) + s(age) + s(lbph) + svi +
                     gleason + s(lcp))
gam_step <- stepGAIC(test_mod, scape = list(lower = ~1, upper = ~s(lcavol) + s(lweight) + s(age) + s(lbph) + svi + gleason + s(lcp)), k = log(nrow(prostate_set1_reduced)),
                     trace = TRUE, direction = "backward")

# Go through the GAM process in the order the step function did to verify
extractAIC(test_mod, k = log(nrow(prostate_set1_reduced)))

extractAIC(gam(lpsa ~ s(lcavol) + s(lweight) + s(age) + svi + gleason + s(lcp), data = prostate_set1_reduced),
           k = log(nrow(prostate_set1_reduced)))

extractAIC(gam(lpsa ~ s(lcavol) + s(lweight) + svi + gleason + s(lcp), data = prostate_set1_reduced),
           k = log(nrow(prostate_set1_reduced)))

extractAIC(gam(lpsa ~ s(lcavol) + s(lweight) + svi + gleason, data = prostate_set1_reduced),
           k = log(nrow(prostate_set1_reduced)))

selected_gam <- gam(data = prostate_set1_reduced, lpsa ~ s(lcavol) + s(lweight) + svi)
extractAIC(selected_gam, k = log(nrow(prostate_set1_reduced)))

# Revisit this to understand how BIC differs through the two methods (stepGAIC vs. extractAIC)
# Selected model feels reasonable but it may be dealing with weird DF values
# Still more work to do here for ii) and iii)

```

# c) Use the best model above but with only linear terms. Discuss differences.

```{r}
# FIX THIS, STILL USING OLD APPROACH


# c) Replace best model with only linear terms using GAM, get bic. Discuss
selected_linear_gam <- gam(data = prostate_set1_reduced, lpsa ~ lcavol + lweight + svi)
extractAIC(selected_linear_gam, k = log(nrow(prostate_set1_reduced)))

# In comparison with the spline model, the BIC of the linear equivalent is actually an improvement. This
# suggests that the flexibility of splines does not add much to this analysis and actually hinders testing
# due to it sucking up more degrees of freedom.

```

# 2. Use the same data as the previous splits
## a) Scale the training data to mean 0, sd 1

```{r}
# 2. Use the same data for the previous splits to achieve the following

# a) Scale the training data to mean 0, sd 1
prostate_train_scaled <- prostate_set1_reduced %>% apply(., MARGIN = 2, scale) %>% as.data.frame()

```

## b)

```{r}
# b) Fit 1, 2, and 3 term PPR regression models to the scaled data. Use a smoothing
# spline instead of the default smoother. Do not use the max.terms argument
ppr_oneterm <- ppr(data = prostate_train_scaled, lpsa ~s(lcavol) + s(lweight) + s(age) + s(lbph) + s(svi) +
      s(gleason) + s(lcp), nterms = 1)

ppr_twoterm <- ppr(data = prostate_train_scaled, lpsa ~s(lcavol) + s(lweight) + s(age) + s(lbph) + s(svi) +
                     s(gleason) + s(lcp), nterms = 2)

ppr_threeterm <- ppr(data = prostate_train_scaled, lpsa ~s(lcavol) + s(lweight) + s(age) + s(lbph) + s(svi) +
                     s(gleason) + s(lcp), nterms = 3)
ppr_oneterm$alpha
```

