---
title: "Lecture_5"
author: "Matthew Reyers"
date: "October 9, 2019"
output: html_document
---

# 1. Plot the time series and comment on the trends.

```{r, warning=FALSE, message=FALSE}
library(datasets)
library(tidyverse)
library(splines)

plot(UKDriverDeaths)

mod_df <- data.frame(deaths = as.matrix(UKDriverDeaths), date = time(UKDriverDeaths))

# 1. Plot. Comment on trends
mod_df %>%
  ggplot(aes(x = date, y = deaths)) +
  geom_point() + geom_line() +
  ggtitle("Deaths by Month of Year in British Auto Accidents") +
  xlab("Month and Year") + ylab("Deaths") +
  theme_bw()

```

The cyclic behavior of driving deaths is evident in the data set. It appears to peak in the Summer months and decline towards the Winter months. In terms of overall trend, there seems to be an immediate reduction in Auto Accident Deaths leading into 1984. I am unsure as to whether this change is real or not as it seems driving deaths are climbing back up towards the end of 1985. Perhaps it is coincidence, though I expect not.

# 2. Fit natural splines with varying degrees of freedom. Comment on the process and which value for degrees of freedom seems to work best.

```{r}

# 2. Fit natural splines with varying DF until fit is good. Comment on the process
basic_mod <- lm(deaths ~ ns(date, df = 1), data = mod_df)
three_mod <- lm(deaths ~ ns(date, df = 3), data = mod_df)
five_mod <- lm(deaths ~ ns(date, df = 5), data = mod_df)
seven_mod <- lm(deaths ~ ns(date, df = 7), data = mod_df)
nine_mod <- lm(deaths ~ ns(date, df = 9), data = mod_df)
sixteen_mod <- lm(deaths ~ ns(date, df = 16), data = mod_df)
thirtytwo_mod <- lm(deaths ~ ns(date, df = 32), data = mod_df)

mod_df %>%
  mutate(one_df = basic_mod$fitted.values,
         three_df = three_mod$fitted.values,
         five_df = five_mod$fitted.values,
         seven_df = seven_mod$fitted.values,
         nine_df = nine_mod$fitted.values,
         sixteen_df = sixteen_mod$fitted.values,
         thirtytwo_df = thirtytwo_mod$fitted.values) %>%
  ggplot(aes(x = date, y = deaths)) +
  geom_point() + geom_line() +
  geom_line(aes(y = one_df, col = '1 DF'), size = 1.5) +
  geom_line(aes(y = three_df, col = '3 DF'), size = 1.5) +
  geom_line(aes(y = five_df, col = '5 DF'), size = 1.5) +
  geom_line(aes(y = seven_df, col = '7 DF'), size = 1.5) +
  geom_line(aes(y = nine_df, col = '9 DF'), size = 1.5) +
  geom_line(aes(y = sixteen_df, col = '16 DF'), size = 1.5) +
  geom_line(aes(y = thirtytwo_df, col = '32 DF'), size = 1.5) +
  ggtitle("Spline Comparison for Driver Deaths") +
  theme_bw()

# A little messy, facet_wrap this in some smart way
mod_df %>%
  mutate(one_df = basic_mod$fitted.values,
         three_df = three_mod$fitted.values,
         five_df = five_mod$fitted.values,
         seven_df = seven_mod$fitted.values,
         nine_df = nine_mod$fitted.values,
         sixteen_df = sixteen_mod$fitted.values,
         thirtytwo_df = thirtytwo_mod$fitted.values) %>%
  gather(DF, Fit, -deaths, -date) %>%
  mutate(DF.f = factor(DF, levels = c('one_df', 'three_df', 'five_df', 'seven_df', 'nine_df', 
                                      'sixteen_df', 'thirtytwo_df'))) %>%
  ggplot(aes(x = date, y = deaths)) +
  geom_point() + geom_line() +
  geom_line(aes(y = Fit), col = 'red', size = 1.5) +
  facet_wrap(~DF.f) +
  theme_bw()

```

It appears to me that the monthly trend is obviously chased with 32 degrees of freedom and slightly chased with 16 degrees of freedom. Somewhere between 5 degrees of freedom and 9 degrees of freedom seems to be about right. From this, I would select 7 degrees of freedom. This selection seems to balance the overall trend with minimal monthly chasing. 

# 3. Fit optimal smoothing splines and compare the results in a) and b).

```{r}

# 3. Fit optimal smoothing spline, compare.
opt_spline <- smooth.spline(mod_df$date, mod_df$deaths, cv = TRUE) # specify LOOCV
opt_spline$y

mod_df %>%
  mutate(seven_df = seven_mod$fitted.values,
         opt_spline = opt_spline$y) %>%
  gather(DF, Fit, -deaths, -date) %>%
  mutate(DF.f = factor(DF, levels = c('seven_df', 'opt_spline'))) %>%
  ggplot(aes(x = date, y = deaths)) +
  geom_point() + geom_line() +
  geom_line(aes(y = Fit), col = 'red', size = 1.5) +
  facet_wrap(~DF.f) +
  theme_bw()

```

# a) How well does the optimal spline meet the objective?

The optimal smoothing spline does a poor job of meeting our objective. Instead it picks a large number of degrees of freedom and chases the monthly trends. 

# b) Are the DF similar to those from the natural splines?

The DF differ greatly between the models. My selected model has only 7 DF while the optimal curve has equivalent degrees of freedom of `r opt_spline$df`.

# 4. Fit loess curves with various spans and degrees until one looks good. Plot the results and discuss.

```{r}

# 4. Fit loess curves with various spans and degrees until one looks good. Plot the results and discuss

loess_1 <- loess(deaths ~ date, data = mod_df, span = 0.75, degree = 1)
loess_2 <- loess(deaths ~ date, data = mod_df, span = 0.75, degree = 2)
loess_3 <- loess(deaths ~ date, data = mod_df, span = 0.5, degree = 1)
loess_4 <- loess(deaths ~ date, data = mod_df, span = 0.5, degree = 2)
loess_5 <- loess(deaths ~ date, data = mod_df, span = 0.25, degree = 1)
loess_6 <- loess(deaths ~ date, data = mod_df, span = 0.25, degree = 2)
loess_7 <- loess(deaths ~ date, data = mod_df, span = 1, degree = 1)
loess_8 <- loess(deaths ~ date, data = mod_df, span = 1, degree = 2)
loess_9 <- loess(deaths ~ date, data = mod_df, span = 1.25, degree = 1)
loess_10 <- loess(deaths ~ date, data = mod_df, span = 1.25, degree = 2)

mod_df %>%
  mutate(loess_1 = loess_1$fitted,
         loess_2 = loess_2$fitted,
         loess_3 = loess_3$fitted,
         loess_4 = loess_4$fitted,
         loess_5 = loess_5$fitted,
         loess_6 = loess_6$fitted,
         loess_7 = loess_7$fitted,
         loess_8 = loess_8$fitted,
         loess_9 = loess_9$fitted,
         loess_10 = loess_10$fitted) %>%
  gather(Fit, Value, -deaths, -date) %>%
  mutate(loess_f = factor(Fit, levels = c('loess_1', 'loess_2', 'loess_3', 'loess_4', 'loess_5',
                                          'loess_6', 'loess_7', 'loess_8', 'loess_9', 'loess_10'))) %>%
  ggplot(aes(x = date, y = deaths)) +
  geom_point() + geom_line() +
  geom_line(aes(y = Value), size = 1.25, col = 'red') +
  ggtitle("Loess comparison to original data") +
  xlab("Date") + ylab("Deaths") +
  facet_wrap(~ loess_f) + theme_bw()

```

Assuming the objective is still to identify the trend and ignore chasing monthly trends, the 4th loess curve above seems to do the most reasonable job. This uses a span of 0.5 and a degree of 2. Using a span of 0.5 implies a fraction of the data will be used in each fit. A degree of 2 gives the most flexible polynomial possible in the loess function. These decisions make sense in the  context of cyclical data such as the deaths dataset.

# a) Repeat with a normal kernel

```{r}

# a) Repeat with a normal kernel

library(KernSmooth)
lp_1 <- locpoly(x=mod_df$date, y=mod_df$deaths, gridsize = dim(mod_df)[1], bandwidth=.75, degree=1)
lp_2 <- locpoly(x=mod_df$date, y=mod_df$deaths, gridsize = dim(mod_df)[1], bandwidth=.75, degree=2)
lp_3 <- locpoly(x=mod_df$date, y=mod_df$deaths, gridsize = dim(mod_df)[1], bandwidth=.5, degree=1)
lp_4 <- locpoly(x=mod_df$date, y=mod_df$deaths, gridsize = dim(mod_df)[1], bandwidth=.5, degree=2)
lp_5 <- locpoly(x=mod_df$date, y=mod_df$deaths, gridsize = dim(mod_df)[1], bandwidth=.25, degree=1)
lp_6 <- locpoly(x=mod_df$date, y=mod_df$deaths, gridsize = dim(mod_df)[1], bandwidth=.25, degree=2)
lp_7 <- locpoly(x=mod_df$date, y=mod_df$deaths, gridsize = dim(mod_df)[1], bandwidth=1, degree=1)
lp_8 <- locpoly(x=mod_df$date, y=mod_df$deaths, gridsize = dim(mod_df)[1], bandwidth=1, degree=2)
lp_9 <- locpoly(x=mod_df$date, y=mod_df$deaths, gridsize = dim(mod_df)[1], bandwidth=1.25, degree=1)
lp_10 <- locpoly(x=mod_df$date, y=mod_df$deaths, gridsize = dim(mod_df)[1], bandwidth=1.25, degree=1)

mod_df %>%
  mutate(lp_1 = lp_1$y,
         lp_2 = lp_2$y,
         lp_3 = lp_3$y,
         lp_4 = lp_4$y,
         lp_5 = lp_5$y,
         lp_6 = lp_6$y,
         lp_7 = lp_7$y,
         lp_8 = lp_8$y,
         lp_9 = lp_9$y,
         lp_10 = lp_10$y) %>%
  gather(Fit, Value, -deaths, -date) %>%
  mutate(Fit_f = factor(Fit, levels= c('lp_1', 'lp_2', 'lp_3', 'lp_4', 'lp_5',
                                       'lp_6', 'lp_7', 'lp_8', 'lp_9', 'lp_10'))) %>%
  ggplot(aes(x = date, y = deaths)) +
  geom_point() + geom_line() +
  geom_line(aes(y = Value), col = 'red', size = 1.25) +
  ggtitle("Locpoly comparison of Parameter Values") +
  xlab("Date") + ylab("Deaths") +
  facet_wrap(~ Fit_f) +
  theme_bw()

```


The plots using a Normal Kernel seem to suggest that lp_1, lp_2, and lp_8 match the trend without excessively chasing errors. I would be comfortable using any of the three. lp_1 uses bandwidth = 0.75,  degree 1. lp_2 uses bandwidth = 0.75, degree 2. And lp_3 uses bandwidth = 1, degree 2.

# b) Try the optimal bandwidth with the normal kernel

```{r}

# b) Also try the optimal bandwidth with the normal kernel
lambda <- dpill(x=mod_df$date, y=mod_df$deaths)
lp.1.opt <- locpoly(x=mod_df$date, y=mod_df$deaths, gridsize = dim(mod_df)[1], bandwidth=lambda, degree=1)
lp.2.opt <- locpoly(x=mod_df$date, y=mod_df$deaths, gridsize = dim(mod_df)[1], bandwidth=lambda, degree=2)

mod_df %>%
  mutate(lp_1 = lp.1.opt$y,
         lp_2 = lp.2.opt$y) %>%
  gather(Fit, Value, -deaths, -date) %>%
  mutate(Fit_f = factor(Fit, levels= c('lp_1', 'lp_2'))) %>%
  ggplot(aes(x = date, y = deaths)) +
  geom_point() + geom_line() +
  geom_line(aes(y = Value), col = 'red', size = 1.25) +
  ggtitle("Locpoly comparison of Parameter Values") +
  xlab("Date") + ylab("Deaths") +
  facet_wrap(~ Fit_f) +
  theme_bw()

```

The optimal lambda approach yields, with either degree 1 or 2, a similar fit to the above selections. Neither overly chases the errors while also capturing the overall trend. If I had to choose, I would take the degree 1 fit (lp_1) here as it is not as heavily pulled by the data points on the boundary.


# 5. Compare the best results from each type of smoother. Is one class of smoothers better for this 
# problem?

Given the cyclic nature of the data, I believe the loess smoother to be a more natural fit to this type of data. 

# TODO: Make some actual numerical comparisons, using rMSE on a prediction set or something of the sort.