---
title: "MattAndMeganHmwk"
author: "Matthew Reyers"
date: "November 28, 2018"
output: html_document
---

a) Install the psych package in R and load the dataset “
bfi
”. Remove the NA cases. Plot the correlation 
matrix. 
```{r}
library(tidyverse)
library(psych)
data(bfi)

summary(bfi)

bfi <- bfi %>% filter(complete.cases(.))
cor(bfi)
```

b) Determine how many factors to extract. Will you use the Kaiser criterion? A scree plot? A goodness-of-
fit 
test?
```{r}
scree(bfi, pc = FALSE)
```
I used a scree plot as I find them easy to interpret. As such, I will be using 4 factors for this work.

c) Factor analysis, decide on appropriate method
- Have to decide whether the data is normal or not to determine method for factor analysis. I did a few checks for histograms on each variable and did not notice any normality. Therefore must proceed with Principle Axis Factoring
```{r}
noRot <- fa(bfi, nfactors = 4, fm = "pa", rotate = "none")
noRot$loadings
Rot <- fa(bfi, nfactors = 4, fm = "pa", rotate = "varimax")
Rot$loadings
```
I used the orthogonality rotation as I found the correlations to be minimal in terms of size, only noticing 1 or 2 values greater than 0.5. Considering there are 25^2 correlations in the matrix I doubt that difference is of importance. 

d) Examine rotated and non-rotated loadings. What does rotation accomplish?
In using the rotation, I found the rotated loadings to have fewer entries in subsequent factors, i.e. more information was captured by the first factor than the remaining 3. This helps to simplify the interpretation of the factors as it is more clear what each factor is explaining.

e) Recommendations for those working in the field
Using the results of the rotated factor analysis, I would say that a limited amount of the variation can be explained by the individual variables in this data set but that the summarization to a level of 4 factors can be of use. We are able to capture roughly 35% of the common variance through this analysis which should lead to more logical and informative fits.


